---
title: "Orthofinder and CAFE analyses"
author: "Turquetti-Moraes, Dayana Kelly"
date: '2024'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r}
library(readr)
library(cogeqc)
library(tidyverse)
library(ape)
```

## Orthofinder analysis to 30 legume species

```{python}

#python orthofinder2.5.2/orthofinder.py -t 24 -a 24 -S diamond_ultra_sens -M msa -T iqtree -f ./proteome_legumes/

```

## Exploring orthofinder results with Cogeqc package

Almeida-Silva F, Van de Peer Y (2022). _cogeqc: Systematic quality
#>   checks on comparative genomics analyses_. R package version 1.3.1,
#>   <https://github.com/almeidasilvaf/cogeqc>.

```{r Orthogroups}

#orthogroups<-read_orthogroups("/proteome_legumes/OrthoFinder/Results_Nov13/Orthogroups/Orthogroups.tsv")

```

##Read Statistics

```{r}

#ortho_stats <- read_orthofinder_stats('/proteome_legumes/OrthoFinder/Results_Nov13/Comparative_Genomics_Statistics/')

```

##Plot results

```{r Tree}

#tree_ortho<-treeio::read.tree("/proteome_legumes/OrthoFinder/Results_Nov13/Species_Tree/ SpeciesTree_rooted_node_labels.txt")

plot_species_tree(tree_ortho)
```

```{r Plot orthofinder statistics}

plot_orthofinder_stats(
  tree_ortho, 
  xlim = c(-0.04, 0.65),
  stats_list = ortho_stats
)

```

```{r Plot orthogroups overlaping}

plot_og_overlap(ortho_stats)

```

## CAFE analysis
Software for Computational Analysis of gene Family Evolution
Fábio K Mendes, Dan Vanderpool, Ben Fulton, Matthew W Hahn, CAFE 5 models variation in evolutionary rates among gene families, Bioinformatics, 2020;, btaa1022, https://doi.org/10.1093/bioinformatics/btaa1022

#input data

```{r gene counts to 30 legume species}

#ogs_gene_count_to_cafe <- read_delim("/CAFE5/CAFE_input/Orthogroups.GeneCount.CAFE.tsv", 
    # delim = "\t", escape_double = FALSE, 
    # trim_ws = TRUE)

```

```{python - ultrametric tree}

# r parameter was inferred from http://www.timetree.org/

#python /orthofinder2.5.2/tools/make_ultrametric.py -r 68 /proteome_legumes/OrthoFinder/Results_Nov13/Species_Tree/SpeciesTree_rooted.txt

```

```{r}
ultrametric_tree <- read.tree("/CAFE5/CAFE_input/SpeciesTree_rooted.txt.ultrametric.tre")
```

#Filtering large orthogroups

```{python}
#python /CAFE5/docs/tutorial/clade_and_size_filter.py -i ogs_gene_count_to_cafe -o filtered_ogs_gene_count_to_cafe -s 2> filtered.log

```

## Filtered orthogroup counts

```{r}

#filtered_ogs_gene_count_to_cafe <- read_delim("/CAFE5/CAFE_input/filtered_ogs_gene_count_to_cafe", 
    #delim = "\t", escape_double = FALSE, 
    #trim_ws = TRUE)
```

##Run CAFE5

```{bash}

for run in {1..51}; do 
cafe5 -i filtered_ogs_gene_count_to_cafe -t ultrametric_tree -k 3 -o run_$run; done
```

## CAFE output

```{r}

# gamma_family_results_signif <- read_delim("/CAFE5/Output_CAFE_k3/run_4/Gamma_family_results_signif.txt", 
#     delim = "\t", escape_double = FALSE, 
#     trim_ws = TRUE)

```

##Plot the results
We used cafeplotter, available at: https://github.com/moshi4/CafePlotter

```{python}

#cafeplotter -i ./run_4/ -o Plots_cafe_run4/ --format pdf

```


